---
title: 'Chapter 7: R Packages'
output:
  html_document:
    df_print: paged
---

# Exercise 1: Installing Packages

**Your Task:**

1. Install the `xgboost` package.  
2. Install `XML` and `mlr3` packages in one line of code by passing in two package names as a single variable.  

**Consider:** You might want to run `?install.packages()` to see what data type it allows for the package names.

```{r}
# Complete the code below and run the cell.

# Install xgboost -- your code here
# Install XML and mlr3 -- your code here
```

**What are these packages?**

`xgboost` is an efficient and scalable machine learning library designed for gradient boosting. It's widely used for supervised learning tasks, especially in structured/tabular data, due to its high performance and accuracy.

The `XML` package in R provides tools to parse and manipulate XML (eXtensible Markup Language) files, commonly used for data exchange between different systems and applications.

`mlr3` is an R package for machine learning in R that provides a unified interface and infrastructure for various machine learning tasks such as data preprocessing, model training, hyperparameter tuning, and evaluation. It's a successor to the popular mlr package and is designed to be more modular and extendable.

#### Test Your Code (Exercise 1)

Do not modify the cell below. Run the cell below to check if you've successfully installed the packages.

```{r}
library(testthat, quietly = TRUE)
test_that("Installing Packages", {
  # Task 1: Check if tidyverse is installed
  expect_true("xgboost" %in% installed.packages()[, "Package"])
  
  # Task 2: Check if ggplot2 and dplyr are installed
  packages_installed <- c("XML", "mlr3")
  expect_true(all(packages_installed %in% installed.packages()[, "Package"]))
})
```

------------------------------------------------------------------------

# Exercise 2: Read a CSV file -- tidyverse

**Your Task:**
Your objective is to read a CSV file using the `tidyverse` package.  
1. Download "sample_data.csv" from this chapter.  
2. If you're working on the UofT JupyterHub, upload the CSV file to your JupyterHub RStudio so that the server has access to the file. If you are working on your own RStudio, make sure the CSV file is in the same folder as this Rmd file.  
3. Use the appropriate function to read in the CSV file. **Do not use any R built-in function to read the file**. Assign the loaded data to a variable named `csv_data`.

Note: Load the entire `tidyverse` package.

```{r}
# Complete the code below and run the cell.

# Read the CSV file using tidyverse function -- your code here

```

#### Test Your Code (Exercise 2)

Do not modify the cell below. Run the cell below to check if you've successfully loaded the csv file.

```{r}
library(testthat, quietly = TRUE)
test_that("Reading a CSV File with readr", {
  # Check if csv_data is a data frame and has rows
  expect_true(is.data.frame(csv_data) & nrow(csv_data) > 0)
})
```

------------------------------------------------------------------------

# Exercise 3: Creating a Scatterplot -- tidyverse

**Your Task:**
Your objective is to create a scatterplot using functions from ggplot2. Use the data you loaded in Exercise 2 (`csv_data`).

1. Create a scatterplot with `Age` on the x-axis and `Score` on the y-axis.  
2. Make sure the plot has a title "Age vs. Score".  
3. Save the plot to a variable named `scatter_plot`.  

**Consider**: Visit this ggplot2 webpage to learn more about customizing the plot: https://ggplot2.tidyverse.org/reference/labs.html

```{r}
# Complete the code below and run the cell.

# Create a scatterplot using ggplot2 -- your code here

```

#### Test Your Code (Exercise 3)

Do not modify the cell below. Run the cell below to check if you've successfully created the scatterplot.

```{r}
library(testthat, quietly = TRUE)
test_that("Scatterplot with ggplot2", {
  # Check if scatter_plot is of ggplot type and has layers (e.g., geom_point)
  expect_true(inherits(scatter_plot, "ggplot") & length(scatter_plot$layers) > 0)
})
```


------------------------------------------------------------------------

# (Optional) Exercise 4: Data Filtering and Summarization -- tidyverse

**Your Task:**
Using the `csv_data` dataset, your objective is to perform some data filtering and summarization.

1. Filter the `csv_data` dataset to only include individuals with an `Age` greater than the **median age** of the dataset.  
2. For these older individuals, compute **both** the average and the standard deviation of the `Score`.  
3. Save the result to a variable named `stats_result`.  

**Consider**: You'll need to use functions you might not have encountered in the chapter. Consult the `dplyr` documentation or help files (e.g., `?summarize` in RStudio) to figure out how to compute the median for filtering and how to calculate the standard deviation within `summarize()`.

```{r}
# Complete the code below and run the cell.

# Filter and summarize data -- your code here

```

#### Test Your Code (Exercise 4)

Do not modify the cell below. Run the cell below to check if you've successfully filtered the data and calculated the required statistics.

```{r}
library(testthat)
test_that("Advanced filtering and summarization", {
  # Check if stats_result contains the correct summaries
  expect_true(all(c("mean_score", "sd_score") %in% names(stats_result)))
  
  # Ensure there's only one row of summary
  expect_equal(nrow(stats_result), 1)
  
  # Ensure the mean_score is a valid number and within the range 0-100 (assuming scores range from 0 to 100)
  expect_true(is.numeric(stats_result$mean_score) & stats_result$mean_score >= 0 & stats_result$mean_score <= 100)
  
  # Ensure the sd_score is a valid number and within a reasonable range
  expect_true(is.numeric(stats_result$sd_score) & stats_result$sd_score >= 0 & stats_result$sd_score <= (max(csv_data$Score) - min(csv_data$Score)))
  
  # Verify that the filtered data is indeed based on the median age
  median_age <- median(csv_data$Age)
  filtered_data <- csv_data[csv_data$Age > median_age, ]
  expect_equal(mean(filtered_data$Score), stats_result$mean_score)
  expect_equal(sd(filtered_data$Score), stats_result$sd_score)
})

```


