---
title: 'Chapter 15: Restructuring Your Data'
output:
  html_document:
    df_print: paged
---

# Getting Started:

## Load Necessary Libraries

Before starting the exercises, ensure you have the necessary libraries loaded:

```{r}
# Load necessary libraries
library(tidyverse)
```

## Load Data

For these exercises, we'll be using the `CO2` dataset from R's built-in datasets. This dataset contains information about carbon dioxide uptake in grass plants under different CO2 concentration levels and treatments.

Load the dataset into your R environment:

```{r}
# Load the CO2 dataset
data(CO2)
```

------------------------------------------------------------------------

# Exercise 1: Using pivot_longer

**Your Task:**

Transform the `CO2` dataset to a longer format where the columns `uptake` and `conc` become values in a new column called `measurement_type`, and their respective values are stored in a new column named `value`. Assign the result to `long_data`.

```{r}
# Complete the code below and run the cell.

# Your code here
long_data <- 
```

#### Test Your Code (Exercise 1)

```{r,purl=FALSE}
library(testthat, quietly = TRUE)
test_that("Using pivot_longer", {
  expect_true("measurement_type" %in% colnames(long_data))
  expect_true("value" %in% colnames(long_data))
})
```

------------------------------------------------------------------------

# Exercise 2: Using pivot_wider

**Your Task:**

Transform the `long_data` you created in Exercise 1 back to a wider format. Ensure that each measurement type (`uptake`, `conc`) becomes its own column again. Assign the result to `wide_data`.

```{r}
# Complete the code below and run the cell.

# Your code here
wide_data <- 
```

#### Test Your Code (Exercise 2)

```{r,purl=FALSE}
library(testthat, quietly = TRUE)
test_that("Using pivot_wider", {
  expect_true(all(c("uptake", "conc") %in% colnames(wide_data)))
})
```

------------------------------------------------------------------------

# Exercise 3: Combined Use of pivot_longer and pivot_wider

**Your Task:**

First, modify the `CO2` dataset by converting all columns except `Plant` into characters. Then, use `pivot_longer` to create a long format with columns for `Plant`, `variable`, and `value`. Following this, use `pivot_wider` to reshape this long data into a wide format where each `variable` becomes a column again, but keep `Plant` as the row identifiers. Assign the final output to `transformed_data`.

```{r}
# Complete the code below and run the cell.

# Your code here
transformed_data <- 
```

#### Test Your Code (Exercise 3)

```{r,purl=FALSE}
library(testthat, quietly = TRUE)

# Original column names from CO2 dataset, except 'Plant'
original_columns <- setdiff(names(CO2), "Plant")

test_that("Combined Use of pivot_longer and pivot_wider", {
  # Test if 'Plant' column is present
  expect_true("Plant" %in% colnames(transformed_data))

  # Test if all other original columns are present in the transformed data
  for (col in original_columns) {
    expect_true(col %in% colnames(transformed_data))
  }
})

```
