# Cheminformatics I: Querrying Databases



<!-- list of PCB congeners (in PDF form) available here: https://www.epa.gov/sites/default/files/2015-09/documents/congenertable.pdf -->

<!-- this list has been compilled to .csv by Diwen Yang, so save a copy to the book and tutorial repo --> 


<!-- recycle code from here: https://github.com/DavidRossHall/lipinski-rule -->

Goal of this chapter: 

- Introduce concept of cheminformatics
- introduce SMILES & encoding chemical information
- introduce chemicals inventories & databases 
- introduce ToxCast data set, and how to query both

R code in this chapter will: 

- Import .csv w/ all PCBs. 
- Get students to access database using `webchem` package
- Using PCB case number, get pub-chem ID, and scrap information on PCBs 
- Plot data to get an idea of what PCB properties are (add mass spec adducts and stuff)

- Would be cool to get toxcast data, but downloading dataset is a pain in the ass. Scrapping might be another option, but this might also suck a lot. 
- Maybe check pub-chem AHR binding results... pubchem has ~ 75 compounds in AHR testing results, which are here: [https://pubchem.ncbi.nlm.nih.gov/bioassay/39064](https://pubchem.ncbi.nlm.nih.gov/bioassay/39064)



- mention use of retticulate to get rdkit package working, much better then the other cheminformatics package

- reticulate get's it's own chapter, along with brief introduction to python...

- cheminformatics chapter II: usin reticulate and rdkit to generate chemically aware tables. 

```{r}
library(tidyverse)
library(webchem)
```

![Generalized PCB structure. The possible positions of chlorine atoms on the benzene rings are denoted by numbers assigned to the carbon atoms. Image from [Wikimedia Commons](Wikimedia Commons. )](images/cheminformatics/pcb_structure.png)


A list of all possible PCB congeners in PCB form can be found on the EPA's website [here](https://www.epa.gov/sites/default/files/2015-09/documents/congenertable.pdf). Fortunately, we've gone through the trouble of transcribing the information into a `.csv` file (using the [`pdftables` package](https://github.com/expersso/pdftables)). To follow along with this tutorial, you can download the data directly from the table below. 

```{r}
pcbs_raw <- read_csv('data/cheminformatics/pcbs_list.csv')

DT::datatable(pcbs_raw,
              extensions = c('Buttons', 'Scroller'),
              rownames = FALSE,
              options = list(
                dom = 'Bfrtip',
                buttons = c('csv')
                )
                       )

```

```{r, eval = FALSE}
# load your local copy of PCB list. 
pcbs_raw <- <PATH TO YOUR LOCAL COPY OF PCB LIST>
```

## Scrapping PubChem for compound information

[Pubchem](https://pubchem.ncbi.nlm.nih.gov/) is the largest collection of chemical information, with over 113 million compounds in the repository. Looking at the entry for [PCB 77](https://pubchem.ncbi.nlm.nih.gov/compound/36187), you'll notice a plethora of information collected from a variety of fields. Another useful element of Pubchem is they also provide predicted values for compounds under the *Chemical and Physical Properties* section. LogK~ow~ (aka *LogP*) is a a useful property for modelling the partitiining and environmental fate of compounds. For PCB 77, which has been extensively studied, PubChem provides the computed LogP value (`XLogP3` = 6.6), and reported values (LogP = 6.72 by Hansch (1979)^[Hansch, C., A. Leo. Substituent Constants for Correlation Analysis in Chemistry and Biology. New York, NY: John Wiley and Sons, 1979., p. 266])

Unsurprisungly, most chemicals exists theoretically, or have not been throughougly studied and characterized. So the property predictions from pubchem are extremely useful. 

```{r, eval = FALSE}

# using CASRN to get pubchem compound ID (CID)
# Running this will take a while as the pubchem stuff is throttled. 

pcbs_cid <- webchem::get_cid(pcbs_raw$casrn)

```

```{r, eval = FALSE}

# specifying properties to extract from pubchem 

propertiesToGet <- c("MolecularFormula", 
                     "MolecularWeight", 
                     "CanonicalSMILES", 
                     "IsomericSMILES",
                     "IUPACName", "XLogP",
                     "ExactMass", 
                     "MonoisotopicMass",
                     "TPSA", 
                     "HBondDonorCount",
                     "HBondAcceptorCount", 
                     "Volume3D")

# using punchem CID to get list of properties. 

pcbs_props <- pc_prop(cid = pcbs_cid$cid,
                      properties = propertiesToGet)
```

Because querrying takes a while, it's best to save your results so you can quickly load them. We'll combine our pubchem search results with our original table, and save the output for further analysis. 

```{r, eval = FALSE}
pcbs_all <- pcbs_raw %>%
  left_join(., pcbs_cid, by = c("casrn" = "query")) %>%
  mutate(cid = as.numeric(cid)) %>%
  left_join(., pcbs_props, by = c("cid" = "CID")) %>%
  relocate(name) %>%
  janitor::clean_names()

write_csv("data/cheminformatics/pcbs_all.csv", x = pcbs_all)
```

## Looking into our querry data. 

```{r}
pcbs <- read_csv("data/cheminformatics/pcbs_all.csv")

pcbs_cl <- pcbs %>%
  mutate(cl_count = str_count(canonical_smiles, "Cl"))
```
```{r}
p <- ggplot(data = pcbs_cl, 
       aes(x = x_log_p, y = cl_count, label = name)) +
  geom_point()

plotly::ggplotly(p)
```


## Notes to backfill into earlier sections

- if you're missing a compound, check the original .csv file. This CASRN numbers can be chagned if you've opened your file in Microsoft Excel or another spread sheet software. 