# Importing data into R

```{r, echo = FALSE, message = FALSE, error = FALSE, warning = FALSE}
library(tidyverse)
```


Unlike *Excel*, you can't copy and paste your data into R (or RStudio). Instead you need to *import* your data into R so you can work with it. This chapter will discuss how your data is stored, and how to import it into R (with some accompanying nuances). 

## How data is stored

While there are a myriad of ways data is stored, notably raw instrument often record results in a proprietary vendor format, the data you're likely to encounter in an undergraduate lab will be in the form of a `.csv` or *comma-separated values* file. As the name implies, values are separated by commas (go ahead and open any `.csv` file in any text editor to observe this). Essentially you can think of each line as a row and commas as separating values into columns, which is exactly how R and *Excel* handle `.csv` files. 


## read_csv

Importing a `.csv` file into R simply requires the `read.csv` or the `read_csv` function from tidyverse. The first variable is the most important as it's the file path. Recall that R, unless specified, uses relative referencing. So in the example below we're importing the `ATR_plastics.csv` from the `data` sub-folder in our project by specifying `"data/ATR_plastics.csv"` and assigning it to the variable `atr_plastics`. Note the inclusion of the file extension.

```{r atr-ex-import, echo=TRUE, message=TRUE}

atr_plastics <- read_csv("data/ATR_plastics.csv")
```
A benefit of using `read_csv` is that it prints out the column specifications with each column's name (how you'll reference it in code) and the column value type. Columns can have different data types, but a data type must be consistent within any given column. Having the columns specifications is a good way to ensure R is correctly reading your data. 

We can also quickly inspect either through the *Environment* pane in *RStudio* or quickly with the `head()` function. Note the column specifications under the column name.

```{r}
head(atr_plastics)
```

Note how the first line of the `ATR_plastics.csv` has been interpreted as columns names (or *headers*) by R. This is common practice, and gives you a handle by which you can manipulate your data. If you did not intend for R to interpret the first row as headers you can suppress this with the additional argument `col_names = FALSE`. 

```{r}
head(read_csv("data/atr_plastics.csv", col_names = FALSE))
```
Note in the example below that since the headers are now considered data, the entire column is interpreted as character values. This will happen if a single non-numeric character is introduced in the column, so beware of typos when recording data! If we wanted to skip rows (i.e. to avoid blank rows at the top of our `.csv`), we can use the `skip = n` to skip n rows:

```{r}
head(read_csv("data/atr_plastics.csv", col_names = FALSE, skip = 1))
```

### Tibbles vs. data frames

Quick eyes will notice the first line outputted above is `# A tibble: 6 x 5`. `tibbles` are a variation of `data.frames` introduced in section one, but built specifically for the `tidyverse` family of packages. While `data.frames` and `tibbles` are often interchangeable, it's important to be aware of the difference in case you do run into a rare conflict. In these situations you can readily transform a `tibble` into a `data.frame` by coercion with the `as.data.frame()` function, and vice-versa with the `as_tibble()` function. 

```{r}

class(as.data.frame(atr_plastics))
  
```


## Importing other data types

There are other functions to import different types of tabular data which all function like `read_csv`, such as `read_tsv` for tab-separate value files (`.tsv`) and `read_excel` and `read_xlsx` from the `readxl` package to import *Excel* files. Note most *Excel* files have probably been formatted for legibility (i.e. merged columns), which can lead to errors when importing into R. If you plan on importing *Excel* files, it's probably best to open them in *Excel* to remove any formatting, and then save as `.csv` for smoother importing into R. 

## Saving data

As you progress with your analysis you may want to save intermediate or final datasets. This is readily accomplished using the `write.csv` (base R) or `write_csv` (tidyverse) functions. Similar rules apply to how we used `read_csv`, but now the second argument specifies the save location and file name, the first argument is which `tibble`/`data.frame` we're saving. Note that R *will not* create a folder this way, so if you're saving to a sub-folder you'll have to make sure it exists or create it yourself. 

```{r, eval=FALSE}
write_csv(atr_plastics, "data/ATRSaveExample.csv")
```

A benefit of `write_csv` is that it will always save in `UTF-8` encoding and ISO8601 time format. This standardization makes it easier to share your `.csv` files with collaborators/yourself.

## Further Reading

See Chapters [10](https://r4ds.had.co.nz/tibbles.html) and [11](https://r4ds.had.co.nz/data-import.html) of *R for Data Science* for some more details on `tibbles` and `read_csv`. 









