# Modelling: Multiple Linear Regressions

In the realm of environmental chemistry, understanding the quantitative relationship between chemical concentrations and corresponding instrumental signals is paramount. While simple linear regression models offer insights into this relationship for individual analytes, the complexity of real-world chemical systems often necessitates more nuanced approaches. 

In this chapter, we delve into the world of multiple linear regressions, where we extend our analysis beyond single analytes to simultaneously model the influence of multiple factors on observed signals. By leveraging the power of multiple linear regression, chemists gain a deeper understanding of the intricate interplay between chemical concentrations and instrumental responses, empowering them to make more informed decisions and extract richer insights from their analytical data. Join us as we embark on this journey to unlock the potential of multiple linear regressions in the realm of analytical chemistry!

## Building Multiple Linear Regressions

### Dataset

The dataset `metal_conc` contains information on metal concentrations measured (parts per million, or ppm) and corresponding signal values (counts per seconds, or cps) for 4 types of metals: Aluminum, Magnesium, Calcium, and Potassium. Take a look at the data before proceeding with our analysis:

```{r}
# Load the necessary libraries
library(readr)
library(dplyr)
library(DT)

# Read the CSV file
metal_conc <- read_csv("data/metal_concentration.csv", show_col_types = FALSE)
DT::datatable(metal_conc)
```

### Regressions

Now, we will build separate linear regression models for each metal to predict signal values based on metal concentrations.

```{r}
# Create a list to store slope and intercept for each metal
metal_results <- list()

# Iterate over each metal
for (metal in colnames(metal_conc)[-1]) {
  # Build linear regression model
  model <- lm(metal_conc[[metal]] ~ metal_conc[["Concentration"]], data = metal_conc)
  
  # Extract slope and intercept
  slope <- coef(model)[2]
  intercept <- coef(model)[1]
  
  # Print slope and intercept
  cat("For", metal, ":\n")
  cat("Slope:", slope, "\n")
  cat("Intercept:", intercept, "\n\n")
  
  # Store slope and intercept in metal_results list
  metal_results[[metal]] <- list(slope = slope, intercept = intercept)
}
```

### Slopes and Intercepts 

We have successfully built linear regression models for each metal, and the slope and intercept values have been stored in the `metal_results` list. This variable is a list containing key-value pairs for each metal, where the metal name serves as the key and the corresponding slope and intercept values are stored as sublists. To visualize the `metal_results` variable, we can display it in a tabular format:

```{r, echo=FALSE}
# Display metal_results variable
metal_results_df <- data.frame(Metal = names(metal_results),
                                Slope = sapply(metal_results, function(x) x$slope),
                                Intercept = sapply(metal_results, function(x) x$intercept))
DT::datatable(metal_results_df)
```


To access the slope and intercept variables for a specific metal from the `metal_results` list, we can use the dollar sign notation (`$`). For instance, to retrieve the slope and intercept for calcium (Ca), we can use the syntax `metal_results$Ca$slope` and `metal_results$Ca$intercept`, respectively. This allows us to directly access the slope and intercept values associated with the metal of interest. The following code snippet illustrates how to access these variables:

```{r}
# Access slope and intercept for calcium (Ca)
slope_ca <- metal_results$Ca$slope
intercept_ca <- metal_results$Ca$intercept

# Print the slope and intercept values
cat("Slope for Ca:", slope_ca, "\n")
cat("Intercept for Ca:", intercept_ca, "\n")
```


## Using Regression Variables for Inference
To calculate the concentration of the metal ions from the provided signal values (in counts per second, cps), we can use the slope and intercept values obtained from the linear regression models. 

Take a look at the following signal values that we don't know the concentrations of:
```{r}
# Read the metal signal data
metal_signals <- read_csv("data/metal_signals.csv", show_col_types = FALSE)
DT::datatable(metal_signals)
```

Let's assume that the linear regression models were built using the concentration of each metal as the independent variable and the signal values as the dependent variable. Here's how we can calculate the concentration for each metal:

```{r}
# Function to calculate concentration from signal values
calculate_concentration <- function(signal, metal_name) {
  # Extract slope and intercept for the specified metal
  slope <- metal_results[[metal_name]]$slope
  intercept <- metal_results[[metal_name]]$intercept
  
  # Calculate concentration using the linear regression equation: concentration = (signal - intercept) / slope
  concentration <- (signal - intercept) / slope
  
  return(concentration)
}

# Apply the calculate_concentration function to each metal column in the dataset
metal_signals$Al_concentration <- calculate_concentration(metal_signals$Al, "Al")
metal_signals$Mg_concentration <- calculate_concentration(metal_signals$Mg, "Mg")
metal_signals$Ca_concentration <- calculate_concentration(metal_signals$Ca, "Ca")
metal_signals$K_concentration <- calculate_concentration(metal_signals$K, "K")

# Print the updated dataset with calculated concentrations (Aluminum as the example)
DT::datatable(metal_signals)

```

Let's take a closer look at one of the analytes' calculated concentrations:

```{r, echo=FALSE}
DT::datatable(metal_signals[, c("Ca", "Ca_concentration")])
```

Do you see anything odd about the calculated concentrations of Calcium in the above table? 

Note that negative concentration values can arise when signal values fall below the intercept of the linear regression model, indicating a deviation from the model's assumptions. This situation typically occurs at lower signal levels, where instrument noise or background signal may influence measurements. Negative concentrations lack meaningful interpretation in chemical analysis and are often considered artifacts of the model's limitations. To address this issue, we could establish a minimum detectable concentration threshold, treating signal values below this threshold as non-detects or zero values. Alternatively, quadratic or logarithmic regression models may better capture non-linear instrument responses, enabling accurate concentration determination across a wider dynamic range. And the great news is, we'll see how to do these non-linear regressions in the next chapter!


## Conclusion

In conclusion, while multiple linear regression serves as a powerful tool for analyzing the relationship between chemical concentrations and instrumental signals, its applicability may be limited by the assumption of linearity. Negative concentration values observed in linear regression analyses highlight the need for robust statistical techniques and critical evaluation of instrument responses. As we move forward into the realm of non-linear regression in the subsequent chapter, we explore alternative modeling approaches that can better accommodate non-linear relationships between variables. By embracing the versatility of non-linear regression techniques, we expand our analytical toolkit, allowing for more accurate and comprehensive analysis of complex chemical systems. Through a combination of linear and non-linear regression methods, chemists can unlock deeper insights into the quantitative relationships governing chemical phenomena, paving the way for advancements in analytical chemistry and beyond.

```{r child='end-of-chapter-exercise-temp.Rmd'}
```