
<!-- This header creates the chapter -->
# R coding basics

Now that you know how to navigate and run code in RStudio, we'll take a look at the basics of R. As we're chemist first, and not computer programmers, we'll try and avoid as much of the nitty-gritty underneath the hood aspects of R. However, a risk of this approach is being unable to understand errors and warnings preventing your code from running. As such, we'll introduce the most important and pertinent aspects of the R language to meet your environmental chemistry needs. 

```{r}
(1000 * pi) / 2

(2 * 3) + (5 * 4)
```

## Variables

Variables in R, like those you've encountered in math class, can only have one value. Unlike the variables in math classes, you can change that value whenever you want. Another way to think about it is to treat variables in R like a box which you use to store a value. Then you can simply open the box somewhere else without having to worry about the hassle of what's inside. 

You can assign the outputs variables using `<-`, as shown below. 

```{r}
x <- 12
x
```
In addition to reading code top to bottom, you often *read it from right to left*. `x <- 12` would be read as "take the value 6.0221e23 and store it into the variable x". The second line of code, `x`, simply returns the value stored inside `x`. You can now use this variable in snippets of code:

```{r}
x
x <- x * 6.022e23
x
```

Remember, we're evaluating from right to left, so the code above is taking the number `6.022e23` and multiplying it by the value of `x`, which is 12 and storing that value back into `x`. Note that variable names are case sensitive, so if your variable is named `x` and you type `X` into the console, R will not be able to print the contents of `x`. Variable names can consist of letters, numbers, dots (`.`) and/or underlines (`_`). Before you create variables of your own. Here are some rules and guidelines for naming variables in R: 

  - **Rules** dictated by R
    - names must begin with a letter or with the dot character. `var` and `.var` are acceptable.
    - Variable names *cannot* start with a number or the `.` character cannot be preceded by  number. `var1` is acceptable, `1var` and `.1var` are not. 
    - Variable names *cannot* contain a space. `var 1` is interpreted as two separate values, `var` and `1`. 
    - Certain words are reserved for R, and cannot be used as variable names. These include, but are not limited to, `if`, `else`, `while`, `function`, `for`, `in`, `next`, `break`, `TRUE`, `FALSE`, `NULL`, `Inf`, `NA`, and `NaN`

Good names for variables are short, sweet, and easy to type while also being somewhat descriptive. For example, let's say you have an air pollution data set. A good name to assign the data set to would be `airPol` or `air_pol`, as these names tell us what is contained in the data set and are easy to type. A bad name for the data set would be `airPollution_NOx_O3_June20_1968`. While this name is much more descriptive than the previous names, it will take you a long time to type, and will become a bit of a nuisance when you have to type it 10+ times to refer to the data set in a single script. Please refer to the [*Style Guide*](http://adv-r.had.co.nz/Style.html) found in *Advanced R* by H. Wickham for more information. 

## Data Types

Data types refer to how data is stored and handled by and in R. You can really get into the weeds on this, but we'll focus on the most common types so you can get started on your work. Firstly, here are the data types you'll likely be working with:

  - **character**: `"a"`, `"howdy"`, `"1"`, is used to represent string values in R. Basically it's text that you'd read. Note the quotation marks and the fact that `"1"`, despite being a number, is stored as a character.
  - **numeric** (real or decimal): `2`, `3.14`, `6.022e23`.
  - **integer**: `2L`, note the 'L' tells R this is an integer. 
  - **logical**: either `TRUE` or `FALSE` 
  
There are some helpful functions to test the data type of a value in R as a frequent source of error and frustration are values stored in the wrong data type. For example: 

```{r, error = TRUE}
x <- "6"

x / 2
```
You'll see the error above frequently, and it's simply saying you're trying to do math on something you can't do math on. You might think if x is 6, why can't I divide it by 2? Let's see what type of data `x` is: 

```{r}
is.numeric(x)   # test if numeric 
is.logical(x)   # test if logical
is.integer(x)   # test if integer 
is.character(x) # test if character
```
So the value of `x` is a character, in other words R treats it as a word, and we can't do math on that (think, how would you divide a word by a number?). So let's convert the data type of `x` to numeric to proceed. 

```{r}
x <- as.numeric(x)
is.numeric(x)
x / 2
```

So we've converted our character string `"6"` to the numerical value `6`. Keep in mind there are other conversion functions which are described [elsewhere](https://www.geeksforgeeks.org/conversion-functions-in-r-programming/), but you can't always convert types. In the above example we could convert a character to numeric because it was ultimately a number, but we couldn't do the same if the value of `x` was `"six"`. 
  
## Data Structures

Data structures refers to how R stores data. Again, it's easy to get lost in the weeds here so we'll focus on the most common and useful data types for your work which will overwhelmingly be data frames. Data frames consist of data stored in rows and columns. If you've ever worked with a spreadsheet (i.e. *Excel*), it's essentially like that with the caveat that *all data stored in a column must be of the same type*. Again, different columns can have different data types, but *within* a column all the data needs to be the same type. R will convert your data otherwise to make it all the same. A common error is a single character in a column of numerical values leading to the entire column to be interpreted as character values. Errors like this most often stem from errors in recording and importing your data so be careful!

Let's import some data to see a data frame:

```{r, message = FALSE}
airPol <- read_csv("data/2018-07-01_60430_Toronto_ON.csv")

# data table so you can browse the data online
DT::datatable(airPol)
```

The above is air quality data measured in downtown Toronto around July 2018. Some of the variables are:

  - `naps`, `city`, `p`, `latitude`, `longitude` to tell you where the data was measured. 
  - `data.time` for when the measurements were taking. 
  - `pollutant` for the chemical measured
  - `concentration` for the measured concentration in parts-per-million (ppm). 

This data is stored **tidy**, which is to say each column is a variable and each row is an observation. So reading the first row, we know that the Toronto 60430 station on 2018-07-01 at midnight measured ambient O~3~ concentrations of 46 ppm. The concept of tidy data is important and is integral to working in R. It's discussed further in [Tidying your data]. For now, it allows us to easily plot our data using the `ggplot2` package (discussed further in [Visualizations]):

```{r}
library(tidyverse)

ggplot(data = airPol, 
       aes(x = date.time, 
           y = concentration, 
           colour = pollutant)) +
  geom_line()
```

### Other data structures

R has several other data structures. They aren't as frequently used, but it's worth being aware of their existence. Other structures include: 

  - **Vectors** contain multiple elements *of the same type*; either numeric, character (text), logical, or integer. Vectors are created using `c()`, which is short for combine. A data frame is just multiple vectors arranged into columns. Some examples of vectors are shown below. 

```{r}
num <- c(1, 2, 3, 4, 5)
num

char <- c("blue", "green", "red")
char

log <- c(T, T, T, F, F, F)
log
```
  - **Lists** are similar to vectors in that they are one dimensional data structures which contain multiple elements. However, lists can contain multiple elements of different types, while vectors only contain a single type of data. You can create lists using `list()`. Some examples of lists are shown below. You can use `str()` to reveal the different components of a list, in a more detailed format than if you were to simply type the assigned name of the list.

```{r}
hi <- list("Hello", c(5,10,15,20), c(T, T, F))
str(hi)
hi
```

There are many freely available resources online which dive more in depth into different data structures in R. If you are interested in learning more about different structures, you can check out the  [*Data structure*](http://adv-r.had.co.nz/Data-structures.html) chapter of *Advanced R* by Hadley Wickham.

## R packages and functions

Functions are bits of code written to execute a specific task. They offer a convenient means to reduce the amount of typing while making code more reliable and readable. Earlier we use the `ggplot` function to create a time series plot of ambient pollutant measurements in Toronto. We could have made a similar plot using functions written in base R, but often other people have done this and created improved functions to meet there's and others needs. Another example of packages is the `tidyverse` suite of packages that are the base of this book as shown in [Installing packages]. 

An oft unappreciated aspect of packages is that they not only contain functions we can use, but documentation. Documentation provides a description of the function (what it does), what arguments it takes, details, and working examples.Often the easiest way to learn how to use a function is to take a working example and change it bit by bit to see how it works etc. To see documentation check the "help" tab in the "outputs" window or type a question mark in front of a functions name: 

```{r}
# Takes you to the help document for the ggplot function
?ggplot
```
You can also write you're own functions. Please see [Programming with R] for additional details. 
  
Now that you're familiar with navigating RStudio and some basic coding building blocks, let's move over to Chapter 3, where we'll review a normal workflow in R.
