## Loops

Repetition is a key component of programming. Loops enable you to execute the same piece of code multiple times, making data processing more efficient. If you have a previous programming experience in any language, these syntax should look familiar.

#### The `for` loop

The `for` loop in R is used to iterate over elements in a vector or list.

```{r}
for (i in 1:5) {
  print(i)
}
```

This will print numbers from 1 to 5. The loop iterates over each element in the sequence `1:5`, setting the value to `i` and executing the code inside the loop.

#### The `while` loop

The `while` loop continues executing as long as a specified condition remains true.

```{r}
count <- 1

while (count <= 5) {
  print(count)
  count <- count + 1
}
```

In this example, the loop will keep printing and incrementing the value of counter until counter is no longer less than or equal to 5.

Remember, be cautious with while loops. If the condition never becomes false, the loop will run indefinitely!



## Writing custom functions in R

Functions form the backbone of most programming languages, and R is no exception. While R provides a rich library of built-in functions, there are times when you'll need to define your own. Custom functions in R allow you to encapsulate a series of commands into a single reusable unit.

**Why write custom functions?**

-   **Reusability**: Once you've written and tested a function, you can use it repeatedly without having to retype or copy-paste the same lines of code.
-   **Maintainability**: Changes can be made in one place (inside the function) rather than at multiple locations where the code might be used.
-   **Clarity**: Well-named functions can make your main code more readable, as they abstract away the complexity.

#### Simple example

```{r}
add_two <- function(a, b) {
  # Add a and b and return the value
  added <- a + b
  return(added)
}
```

-   `add_two` is the custom name of the function.
-   `a` and `b` are the inputs to the function. You can name them `x` and `y`, or anything else, as long as they are kept consistent throughout the function definition. You can also change the number of inputs for your function.
-   The `return` statement specifies the output of your function. If omitted, the function will return the result of the last expression evaluated.

You can run this function with various choices of `a` and `b`:

```{r}
add_two(1, 3)
add_two(20, 35.5)
```

At this time, we shall keep it simple. You will eventually see more complex usages of custom functions.