---
title: 'Chapter 5: Workflows for R Coding'
output:
  html_document:
    df_print: paged
---

# Exercise 1: Saving R Objects

You have a numeric vector `my_vector` with the values 10, 20, 30, 40, and 50.

**Your Task:**

1. Save the `my_vector` to a CSV file named "my_vector.csv" in the working directory.  
2. Read the "my_vector.csv" (do not assign it to a variable).  
3. When you run the code, what data type is the output? Is this what you expected?  

**Consider:** What does it mean by "working directory," and how do we specify that in our code?

```{r}
# Complete the code and run the cell.
my_vector <- c(10, 20, 30, 40, 50)

# Your code here to save my_vector to "my_vector.csv"

# Your code here to read "my_vector.csv".

```

**Note:** When you use `read.csv()` to read data from a CSV file, the output is a data frame, which is like a table with rows and columns. Even though you might expect a list, a data frame is a way for R to make sure your data stays structured and easy to work with. If you need just the data as a list, you can easily get it by using `read.csv(...)[, 1]`, which takes the first column and turns it into a list. 

#### Test Your Code (Exercise 1)

Do not modify the cell below. Run the cell below to check if you've defined and assigned the variables correctly.

```{r}
# Load testthat package
library(testthat)

# Unittest for Exercise 1
test_that("Exercise 1: Saving R Objects (CSV file)", {
  # Check if "my_vector.csv" exists
  expect_true("my_vector.csv" %in% list.files())
})
```

------------------------------------------------------------------------

# Exercise 2: Code Troubleshooting and Readability

Modify the following R script.  

**Your Task:**

1. There is a line that will result in an error. What is wrong with it, and what error message do you expect to see?  
2. Fix the incorrect line so that the cell will run without an error.  
3. Add appropriate comments to enhance readability. Use `#` to write a comment in each line.  

**Consider:** As a hint, visit [this R documentation](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/numeric).

```{r}
# Modify and run this cell.

a <- 5
b <- "10"
result <- a + b
print(result)
```

#### Test Your Code (Exercise 2)

Do not modify the cell below. Run the cell below to check if you've defined and assigned the variables correctly.  

```{r}
# Load testthat package
library(testthat)

# Unittest for Exercise 2
test_that("Exercise 2: Code Troubleshooting and Readability", {
  expect_true(result == 15)
})
```
