# Ggplot Basic Visualizations 

`ggplot2` is loaded by default with the `tidyverse` suite of packages. Let's revisit our spectroscopy data we encountered in [Tidying Your Data]:


```{r, warning=FALSE}
library(tidyverse)
atr_data <- read_csv("data/ATR_plastics.csv")

# First 50 rows of data
DT::datatable(head(atr_data, 50))
```

## Building plots ups

The `gg` in `ggplot2` stands for the *grammar of graphics*, [@Wickham2009] and it's a way to break down graphics (plots) into small pieces that can be discussed (hence grammar). We'll take a look at this grammar via `geoms` (what kind of plot), `aes` (aesthetic choices), etc. For now, understand that this means we need to build up graphics/plots piece-by-piece and layer-by-layer. This extends beyond code to how we code. Plot often, and discard the useless ones. Take the time to pretty up your plot *after* you're satisfied with the underlying data.

## Basic plotting

`ggplot2` uses `geoms` to specify what type of plot to create. Different plots are used to tell different stories and have different strengths and weakness. We'll explore these more in [Visualizations for Env Chem], but for now we'll focus on `geom_point()`, which simply plots data as points on the 2-D plane. In other words, a scatter plot.

Let's plot our tidied `atr_data` data:

```{r}
ggplot(data = atr_data, 
       aes(x = wavenumber, y = absorbance)) +
  geom_point()
```


Let's ignore the plot for now and look at our code below:

1. `ggplot()` creates a *ggplot object* that contains specifications for the plot:
  - We want to plot data from our `atr_data` dataset (`data = atr_data`).
  - We specified our *aesthetic mappings* via `aes()`, which communicates how we want the plot to look. In this case, we've specified which values from `atr_data` are the x-axis values (`x = wavenumber`) and y-axis values (`y = absorbance`).
2. We then add the `geom_point()` layer to create a scatter plot of (x,y) points.

Now let's look at our result. What we see is a point for every recorded absorbance measurements from our ATR analysis. We can clearly see the spectra of the different plastics in our data, however they're all coloured the same. This is because we've only specified the x and y values. As far as `ggplot()` is concerned, these are the only values that matter, but we know different.

Fortunately you can pass multiple variables to different `aes()` options to enhance our plot. For instance, we can pass the `sample` variable, which specifies which sample a spectrum originates from, to the `colour` option:

```{r}

ggplot(data = atr_data, 
       aes(x = wavenumber, 
           y = absorbance, 
           colour = sample)) +
  geom_point()

```


Now we have a legend which clearly specifies which points are associated with each sample. But now the points are too large, potentially masking certain peaks. We can adjust the size of each point as follows:

```{r}
ggplot(data = atr_data, 
       aes(x = wavenumber, 
           y = absorbance, 
           colour = sample)) +
  geom_point(size = 0.5)
```


We specified `size = 0.5` in the `geom_point()` call because it's a constant. We can map `size` to any continuous variable, such as the absorbance:

```{r}
ggplot(data = atr_data, 
       aes(x = wavenumber, 
           y = absorbance, 
           colour = sample,
           size = absorbance)) +
  geom_point()
```


Sometimes this makes sense (i.e. a *bubble chart*) but for our example, having the size of the points increase as the absorbance increases doesn't provide any new information (it actually clutters our plot).

## Changing plot labels

By default `ggplot` uses the header of the columns you passed for the `x` and `y` `aes()` options. Because headers are written for code they're often poor label titles for plots. We can specify new labels and plot titles as follows:

```{r}
ggplot(data = atr_data, 
       aes(x = wavenumber, 
           y = absorbance, 
           colour = sample)) +
  geom_point() +
  labs(title = "ATR Spetra",
       subtitle = "Courtesy of CHM317 student data",
       x = "Wavenumber, cm^-1",
       y = "Absorbance (arbitrary units)",
       caption = "hi mom",
       colour = "Plastic")
  
```


Note how we changed the title of the legend with `colour = "Plastics"`. This is because the legend is generated from our colour aesthetic (`aes(..., colour = sample)`). If our legend was based off of the size aesthetic, we would use `size = "New Title"` to change the title for the size legend. 

## Small Multiples

Sometimes your plots become overwhelming, a phenomena called overplotting, which prevent your from comparing graphs or charts. A popular solution is [*small multiples*](https://en.wikipedia.org/wiki/Small_multiple), a series of similar plots using the same scale and axes. This is readily accomplished in R using [`facet_grid()`](https://ggplot2-book.org/facet.html) (which creates a 2-D grid ) or [`facet_wrap()`](https://ggplot2-book.org/facet.html) (a single 1d ribbon wrapped into 2D space). You simply specify which variable you want to differentiate your plots, for us it's `sample`: 

```{r}
ggplot(data = atr_data, 
       aes(x = wavenumber, 
           y = absorbance)) +
  geom_point() +
  facet_wrap(~sample)
```


Note the use of the tilde (`~`) in `facet_wrap(~sample)`; in this situation, it's shorthand telling `facet_wrap()` to make small multiples off of the sample variable. 

## Plotting subsets of data

Often you won't want to plot everything in your dataset. Rather, you'll want to plot a specific chemical, city, location, etc. To that end you want to plot a *subset* of your data. There are a couple of ways to handle this. 

You can subset your data on the fly using `subset()`. This way allows you to specify based off of [Logical operators] as such: 

```{r}
ggplot(data = subset(atr_data, sample == "EPDM"),
       aes(x = wavenumber, 
           y = absorbance)) +
  geom_point()
```

or 

```{r}
ggplot(data = subset(atr_data, wavenumber >= 2500),
       aes(x = wavenumber, 
           y = absorbance)) +
  geom_point()
```

Another approach is to use `filter()` and pipe to `ggplot()`:

```{r}
atr_data %>%
  filter((sample != "EPDM" & wavenumber <= 2500) | (sample == "EPDM" & wavenumber >= 3500)) %>%
  ggplot(aes(x = wavenumber, y = absorbance, colour = sample)) +
  geom_point()
```

There are pros and cons to either approach. `subset()` on the fly is best for simple task, like plotting a single city, whereas the piping approach is best for more complex sorting. 

```{r child='src/common/end-of-chapter-exercise.Rmd'}
```
