---
title: 'Chapter 12: Transformation: Data Manipulation'
output:
  html_document:
    df_print: paged
---

# Getting Started:

**IMPORTANT**: Chapter 11 exercise is a prerequisite to this exercise. If you haven't done so yet, please complete chapter 11 exercises first.

Before diving into the exercises, follow the instructions below to download a dataset from the Exploring Air Quality Data website.

-   Visit the [Exploring Air Quality Data](https://uoft-chem.shinyapps.io/Air_Quality_App/) and go to `My Data` tab.
-   Enter your student number, and then download the CSV file.
-   Rename the downloaded file to `data.csv`.
-   Upload the file to your RStudio.

## Load Necessary Libraries

Before starting the exercises, ensure you have the necessary libraries loaded:

```{r}
# Load necessary libraries
library(tidyverse)
```

Use the `read_csv()` function from the `readr`(`tidyverse`) package to import your Air Quality dataset, and assign the imported data to a variable named `data`.

```{r}
# Complete the code below and run the cell.

# Import data from CSV file using read_csv -- your code here
data <- 
```

**Note**:
For all the exercises in this notebook, remember that while the pipe operator enables seamless chaining of transformations, it's entirely okay to split your code into multiple steps or variables if it helps clarity. It's essential to prioritize intuition and readability in your code.

------------------------------------------------------------------------

# Exercise 1: Ox

In this exercise, we will work with Ox ($O_X$), also known as "odd oxygen".
Before you begin, you are recommended to watch this introductory video on odd oxygen: https://play.library.utoronto.ca/watch/1f19e6aa27d9c50e551ee8848292fe7c

As mentioned in the video, atmospheric concentrations of NO2 and O3 are related to one another. This relationship is so intimate that reporting the concentration of one without the other is of limited utility. To embrace this relationship, atmospheric chemists have defined the term "odd oxygen" or Ox as the sum of the concentrations of NO2 and O3.

**Your Task:**

Transform the `data` based on the following instructions:

1. Create a new column called `Ox` by adding `O3` and `NO2`, and update the `data`.
2. Calculate the mean, median, and standard deviation of `Ox`. Store the results in columns named `mean_Ox`, `median_Ox`, and `std_dev_Ox` respectively using `summarise()`.
3. Assign the result to `Ox_summary`.

```{r}
# Complete the code below and run the cell.


data <- # Your code to create the 'Ox' column here.



Ox_summary <- # Your code to summarize 'Ox' here.
```

#### Test Your Code (Exercise 1)

Do not modify the cell below.

```{r,purl=FALSE}
library(testthat, quietly = TRUE)
test_that("Ox", {
  
  # Check if Ox_summary exists
  expect_true(exists("Ox_summary"))
  
  # Check if Ox_summary has the right columns
  expect_equal(names(Ox_summary), c("mean_Ox", "median_Ox", "std_dev_Ox"))
  
  # Check values are numbers (not NaN or NA)
  expect_true(!is.na(Ox_summary$mean_Ox))
  expect_true(!is.na(Ox_summary$median_Ox))
  expect_true(!is.na(Ox_summary$std_dev_Ox))
  
  # Add more tests as necessary
})

```

------------------------------------------------------------------------

# Exercise 2: Make Data Long and Remove Erroneous Data

At the end of Chapter 10 exercise, you learned that there are some bad data in our dataset.

Similar to Chapter 11, we will transform the `data` to a longer format where the headers `NO2` and `O3` become values in a new column called `analyte`, and their respective values fill a new column called `concentration`. We assign the result to a variable named `longer_data`. You should not modify the provided code related to this transformation. The `pivot_longer()` function is covered in Chapter 15, where you will have the opportunity to learn and practice it in detail.

Here’s the code for this transformation:

```{r}
# Transform data to a longer format
longer_data <- data %>%
  pivot_longer(cols = c(NO2, O3), names_to = "analyte", values_to = "concentration")
```

**Your Task:**

1. Visualize the `Time`(x-axis) vs. `concentration`(y-axis) values as a scatterplot with `longer_data`.

```{r}
# Your code for the scatterplot of Time vs concentration.

```

2. Now, let's ensure the reliability of your data by removing potential measurement errors. Filter out any rows in data where the `concentration` values are less than 0. Update `longer_data` with the filter data.

```{r}
# Your code to filter out rows with concentration values less than 0 goes here.
longer_data <- 
```

4. Let's visualize the updated `longer_data` to confirm that erroneous data are indeed removed.

```{r}
# Your code for the scatterplot of Time vs concentration.

```

Compare this plot with your first scatterplot. Can you spot the differences?

#### Test Your Code (Exercise 2)

```{r,purl=FALSE}
library(testthat, quietly = TRUE)
test_that("Data Cleaning Test", {
  expect_true("analyte" %in% colnames(longer_data))
  expect_true("concentration" %in% colnames(longer_data))
  # Ensure no concentration values are less than 0
  expect_true(all(longer_data$concentration >= 0))

})
```

------------------------------------------------------------------------

# Exercise 3: The Ozone Watch

**Your Task:**

Transform the `data` to only include rows where `O3` values are greater than 30 and then arrange the results by `Time` in descending order. Assign the result to `data_high_ozone`.

```{r}
# Complete the code below and run the cell.

data_high_ozone <- # Your code here
```

#### Test Your Code (Exercise 3)

Do not modify the cell below.

```{r,purl=FALSE}
library(testthat, quietly = TRUE)
is_ordered_desc <- function(vec) {
  all(diff(vec) <= 0)
}
test_that("The Ozone Watch", {
  expect_true(all(data_high_ozone$O3 > 30))
  expect_true(is_ordered_desc(data_high_ozone$Time))
})
```

------------------------------------------------------------------------

# Exercise 4: High Alert Days

**Your Task:**

Transform the `data` based on the following instructions:

1. Create a new column called `O3_Level` where `O3` values above 30 are labeled as "High" and others as "Low".
2. Filter the data to only have rows with "High" `O3_Level` values.
3. Assign the result to `high_alert_data`.

```{r}
# Complete the code below and run the cell.

high_alert_data <- # Your code here
```

#### Test Your Code (Exercise 4)

Do not modify the cell below.

```{r,purl=FALSE}
library(testthat, quietly = TRUE)
test_that("High Alert Days", {
  expect_equal(length(colnames(high_alert_data)), 6)
  expect_true(all(high_alert_data$O3_Level == "High"))
})
```

------------------------------------------------------------------------

# (Optional) Exercise 5: Analyzing Extreme Days

**Your Task:**

Transform the `data` based on the following instructions:

1. Filter the dataset for days where `NO2` is above the 75th percentile.
2. For this subset, compute a new column `O3_Difference` that represents the difference between each `O3` value and the average `O3` value of the subset.
3. Identify and filter only the rows where `O3_Difference` is above the 90th percentile of its distribution in this subset. If you don't know how to get the 90th percentile value, check out the `quantile()` function.
4. Finally, select only the columns `NAPS`, `Time`, and `O3_Difference`.
5. Assign the result to `extreme_days_data`.

```{r}
# Complete the code below and run the cell.

# Feel free to use this variable ready for you. 
percentile_75_NO2 <- quantile(data$NO2, 0.75)

extreme_days_data <- # Your code here

```

#### Test Your Code (Exercise 5)

Do not modify the cell below.

```{r,purl=FALSE}
library(testthat, quietly = TRUE)

test_that("Analysis of extreme days test", {
  
  # Thresholds
  percentile_75_NO2 <- quantile(data$NO2, 0.75)
  avg_O3_filtered <- mean(data[data$NO2 > percentile_75_NO2,]$O3)
  percentile_90_difference <- quantile(data[data$NO2 > percentile_75_NO2,]$O3 - avg_O3_filtered, 0.9)
  
  # Check 1: Filtering by O3_Difference
  expect_true(all(extreme_days_data$O3_Difference > percentile_90_difference))
  
  # Check 2: Column names
  expected_cols <- c("NAPS", "Time", "O3_Difference")
  expect_equal(colnames(extreme_days_data), expected_cols)
  
})

```
