
# (PART\*) Part 4: Visualization in R {-}

# Visualizations for Env Chem

Visualizations have always been an important part of data science and chemistry. Good graphics illuminate trends and patterns you may have otherwise missed and allow us to quickly inspect thousands of values. R via the `ggplot2` package is one of, if not the premier, data visualization language available. This chapter will formally introduce the `ggplot2` package, explain a bit of the logic undergirding its operation, and give you some quick examples of how it works. Afterwards we'll delve deeper into specific visualizations you'll use and encounter in your studies culminating in preparing your plots for publication.


We've already encountered and produced several types of plots to visualize our data. We've also gone over the theory and basic operations of `ggplot()` in the [ggplot basic visualizations] section. Now, we'll expand on these and explicitly walk through the most common data visualization methods you'll encounter in the field of environmental chemistry. Additionally, we'll learn how to get your plots ready for publication.

The plots we'll be covering include:

-   [Bar Chart]
-   [Box Plots]
-   [Histograms]
-   [Scatter Plots]
-   [Interactive Plots]
-   [Plotting for publication]

These are only a smattering of the possible data visualizations you can perform in R. We're focusing on them because of their ubiquity in our field, but they often won't be the ideal visualizations you need to communicate *your story*. We highly recommend you check out the following resources. Not only are they a great source of inspiration, they provide example code to get you up and running. We consult them regularly.

-   [**Data to viz**](https://www.data-to-viz.com/) which features a decision tree to help you decide on what plot would serve you best.
-   [**ggplot2 extensions gallery**](https://exts.ggplot2.tidyverse.org/gallery/) which is the best repository to the plethora of `ggplot2()` extensions. If you need a specialized plot, check here. Odds are someone has a solution to your problem. Some great extensions include [`ggrepel`](https://github.com/slowkow/ggrepel) for easy labelling of points; [`ggpmisc`](https://docs.r4photobiology.info/ggpmisc/) for statistical annotations; and [`ggpubr`](https://rpkgs.datanovia.com/ggpubr/) for publication ready plots, group wise comparisons, and annotation of statistical significance.
-   [**The R Graph Gallery**](https://www.r-graph-gallery.com/index.html) contains hundreds of charts made with R. While it's not as easy to navigate as *Data to viz*, it does contain many more examples; it is definitely worth exploring.

## Discrete vs. Continuous variables 

The type of plots available to you, and how they display, are dependent on the type of data. Namely, whether your data is *discrete* (i.e. can only take particular values) or *continuous* (is not restricted to defined separate values, but can occupy any value over a continuous range). So a variable consisting of cities would be discrete, whereas a variable like concentration of a chemical would be continuous. You can treat numeric data as categorical if you so choose. Understanding the difference between discrete and continuous data will shape how you plot your data.

## Prerequisites 

Additionally, for this section we'll mostly be using the `atlNO2` and `sumAtl` datasets we created in the [Summarizing data] chapter. Now that we've seen how to turn data into a longer format, we'll show you how `atlNO2` was tidied from the raw (wide format) data. This `atlNO2` is identical to the previous `atlNO2` we saw.

```{r}
atlNO2 <- read_csv("data/2018hourlyNO2_Atl_wide.csv", skip = 7, na =c("-999")) %>%
  rename_with(~tolower(gsub("/.*", "", .x))) %>%
  pivot_longer(cols = starts_with("h"), 
               names_prefix = "h", 
               names_to = "hour", 
               names_transform = list(hour = as.numeric),
               values_to = "conc", 
               values_transform = list(conc = as.numeric),
               values_drop_na = TRUE) 

sumAtl <- atlNO2 %>%
  group_by(p, city) %>%
  summarize(mean = mean(conc),
            sd = sd(conc),
            median = median(conc),
            min = min(conc),
            max = max(conc))
```


```{r child='src/common/end-of-chapter-exercise.Rmd'}
```
