# Customizing your plots

Up until now we haven't paid much attention to the explicit aesthetics of plots beyond what we needed for our exploratory analysis. However, many journals, publications, instructors, etc. will want plots to adhere to certain aesthetic standards. There are scores of options to play with, so we recommend you consult the [ggplot2 Cheat Sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/data-visualization.pdf).

## Interactive Plots

Ultimately your visualizations will be printed to a static PDF document, but in the interim having an interactive plot can be helpful for data exploration. The `plotly` package magically makes most `ggplots` interactive with a simple command. Here's an example with our Toronto air quality data:

```{r}
torontoAir <- read_csv("data/2018-01-01_60430_Toronto_ON.csv")

torPlot2 <- ggplot(data = torontoAir,
                   aes(x = date.time,
                       y = concentration,
                       colour = pollutant)) +
  geom_line()

plotly::ggplotly(torPlot2)
```

This is also super useful when surveying spectroscopy data, although the large number of points in those datasets can take a while to render into interactive plotly plots.

## Plot Themes

Overall themes can be applied to ggplot. The simple and minimalist `theme_classic()` is satisfactory for most submissions, but you can peruse the available themes in ggplot [here](https://ggplot2.tidyverse.org/reference/ggtheme.html) or you can explore many more themes in the [`ggthemes` package](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/).

```{r}
# generating example plot to modify 
p <- ggplot(data = torontoAir, 
            aes(x = date.time,
                y = concentration,
                colour = pollutant)) +
  geom_point() 

# Default theme
default <- p + labs(title = "Default theme")

# Classic theme
classic <- p + 
  theme_classic() +
  labs(title = "Classic theme")
 
# Arranging into grid
gridExtra::grid.arrange(default, classic, ncol = 2)
```

## Legends

You can specify the position of the legend under the `theme()` option as such:

```{r}
bottom <- p + theme(legend.position = "bottom")

inside <- p + theme(legend.position = "inside",
                    legend.position.inside = c(.5, .5))

gridExtra::grid.arrange(bottom, inside, ncol = 2)
```

Other legend positions include: `"left"`, `"right"`, `"bottom"`, `"top"` and `"none"` (remove legend entirely). Use `legend.position.inside` with a two-element numeric vector to specify the location. For example, use `c(0.95, 0.95)` for inside the top-right corner and `c(0.05, 0.05)` for inside the bottom right corner.

## Modifying labels

The labels generated for the plots are derived from the variable names passed along to the `ggplot()` function. Consequently, variable names that are easy to code become ugly labels on the plot. You can modify labels using the `labs()` function. How to use the `labs()` function was also discussed in the [Bar Charts] section of the [Common Types of Graphs] chapter. Note that in this example we changed the legend's title by specifying what `aes()` option we used to create the legend; in the example below it's `colour`.

```{r}
p + labs(title = "Toronto Air quality",
         subtitle = "from Jan 1st to 8th, 2018", 
         x = "Date",
         y = "Concentration (ppb)",
         colour = "Pollutant")
```

## Modifying Axis

We've already talked about labeling axis titles in [Modifying labels], and adding marginal plots in [Scatter plots]. So we'll just briefly touch upon some simple axis modifications.

### Transforming axis

Transformations are largely related to *continuous* data, and are done using `scale_y_continuous()` or `scale_x_continuous()` functions. For example to scale the y-axis of our plot we'd do the following:

```{r}
p +
  scale_y_continuous(trans = "log10") +
  labs(y = "Log10(concentration)")
```

Other useful transformations include "log2" for base-2 logs, "date" for dates, and "hms" for time. The latter two are useful if R hasn't correctly interpreted your dataset. The data type for the `data.time` column of our dataset was correctly interpreted during our initial importation using `read_csv()`. Hooray for doing it right the first time.

### Axis limits

The limits of plots created with `ggplot()` are automatically assigned, but you can override these using the `lims()` function. For example we can specify the limits of our example plot to show from 0 to 100 ppb:

```{r}
p + lims(y = c(0, 100))
```


### Axis ticks/labels

Sometimes when you are plotting, the length of the axis labels is unreadable. This is often the case with categorical data, such as the names of cities like we've encountered earlier. We addressed this earlier in [Bar charts] by rotating the plot 90$^\circ$ with the `coord_flip()` function. This is often the best solution as it's how we read English. Another solution is to rotate the axis labels themselves:

```{r}
basePlot <- ggplot(data = filter(sumAtl, p == "NL"),
                   aes(x = city, 
                       y = mean)) +
  geom_col()

default <- basePlot +
  labs(title = "default plot")

flip <- basePlot +
  coord_flip() +
  labs(title = "coord_flip()")

rotated <- basePlot +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(title = "element_text(angle = 45)")

rotatedHJust <- basePlot +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "element_text(angle = 45, hjust = 1)")


gridExtra::grid.arrange(default, flip, rotated, rotatedHJust, ncol = 2, nrow = 2)
```


## Arranging plots

We talked about how facets can be used to generate multiple plots from a dataset (small multiples), but sometimes you want to combine two or more *different* plots together. There are a couple of ways, but we've been using `grid.arrange()` from the `gridExtra` package (as demonstrated above). You can read up on [`gridExtra` here](https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html). There is also the [`ggarrange` function](https://rpkgs.datanovia.com/ggpubr/reference/ggarrange.html) from the `ggpubr` package which, amongst other things, can easily create shared legends between plots.

```{r}
colchart <- ggplot(data = sumAtl,
                   aes(x = fct_reorder(city, mean),
                       y = mean, 
                       fill = p)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean - sd, 
                    ymax = mean + sd)) +
  coord_flip() 

boxplot <- ggplot(data = atlNO2, 
                  aes(x = city, 
                      y = conc,
                      fill = p)) +
  geom_boxplot() +
  coord_flip()

boxplot

ggpubr::ggarrange(colchart, 
                  boxplot, 
                  ncol = 2, 
                  nrow = 1,
                  labels = c("A", "B"),
                  common.legend = TRUE, 
                  legend = "bottom")
```



```{r child='src/common/end-of-chapter-exercise.Rmd'}
```
