---
title: 'Chapter 10: Tidying Your Data'
output:
  html_document:
    df_print: paged
---

# Getting Started:

## Import Data

Before diving into the exercises, follow the instructions below to download a dataset from the Exploring Air Quality Data website.

-   Visit the [Exploring Air Quality Data](https://uoft-chem.shinyapps.io/Air_Quality_App/) and go to `My Data` tab.
-   Enter your student number, and then download the CSV file.
-   Upload the file to your RStudio. 

Use the `read_csv()` function from the `readr`(`tidyverse`) package to import your Air Quality dataset, and assign the imported data to a variable named `data`.

```{r}
# Complete the code below and run the cell.

# Import data from CSV file using read_csv -- your code here
data <- 
```


## Visualization Before Data Cleaning

To understand the nature of our data, we'll start by visualizing the `O3` vs. `NO2` values. By plotting these values, you'll get a clearer idea of any outliers or potentially erroneous data points. 

```{r}
# Your code for the scatterplot of O3 vs. NO2 before cleaning goes here.
```

Observe the plot. Do you notice any data points that seem unusual or don't fit the overall pattern, especially those with negative values? These could be indicative of measurement errors.


------------------------------------------------------------------------

# Exercise 1: Remove Erroneous Data

Now that you've noticed some erroneous data in your dataset, you need to ensure the reliability of your data by removing potential measurement errors.

**Your Task:** Filter out any rows in data where the `O3` or `NO2` values are less than 0. Update `data` with the filter data.

```{r}
# Your code to filter out rows with O3 or NO2 values less than 0 goes here.
data <- 
```


#### Test Your Code (Exercise 1)
```{r}
library(testthat, quietly = TRUE)
test_that("Data Cleaning Test", {
  
  # Ensure no O3 or NO2 values are less than 0
  expect_true(all(data$O3 >= 0))
  expect_true(all(data$NO2 >= 0))
})
```


**Your Task:** Now that you cleaned your data, re-visualize the `O3` and `NO2` values to confirm the effect of your data cleaning.

```{r}
# Your code for the scatterplot of O3 vs. NO2 before cleaning goes here.
```

Compare this plot with your initial scatterplot. Can you spot the differences? 

------------------------------------------------------------------------

# Exercise 2: Making Data Longer

**Your Task:**

Transform the `data` to a longer format where the headers `NO2` and `O3` become values in a new column called `analyte`, and their respective values fill a new column called `concentration`. Assign the result to `longer_data`.

```{r}
# Complete the code below and run the cell.

# Your code here
longer_data <- 
```

#### Test Your Code (Exercise 2)

Do not modify the cell below. 

```{r}
library(testthat, quietly = TRUE)
test_that("Making data longer", {
  expect_true("analyte" %in% colnames(longer_data))
  expect_true("concentration" %in% colnames(longer_data))
})

```


------------------------------------------------------------------------

# Exercise 3: Separate Date and Time

**Your Task:**
Use the `longer_data` to separate the `Time` column into distinct `year`, `month`, `day`, and `time` columns. Assign the result to `date_time_data`. Consider using the pipe operator (`%>%`) to break down into multiple separation steps.

```{r}
# Complete the code below and run the cell.

# Your code here
date_time_data <- 
```
You might notice that there are NA values in the `Hour` column. Why do you think this is happening? It's okay to leave this for now, but feel free to figure out how to resolve this issue.


#### Test Your Code (Exercise 3)

Do not modify the cell below. 

```{r}
library(testthat, quietly = TRUE)
test_that("Separate date and time", {
  expect_true(all(c("year", "month", "day", "time") %in% colnames(date_time_data)))
})
```


------------------------------------------------------------------------

# Exercise 4: Uniting Columns

**Your Task:**

Using the `date_time_data` from the previous exercise, unite the `year`, `month`, and `day` columns into a single `date` column. Use a space as the separator. Assign the result to a variable named `united_data`.

```{r}
# Complete the code below and run the cell.

# Your code here
united_data <- 
```


#### Test Your Code (Exercise 4)

Do not modify the cell below. 
```{r}
library(testthat, quietly = TRUE)
test_that("Uniting Columns", {
  expect_true("date" %in% colnames(united_data))
})
```


------------------------------------------------------------------------

# Exercise 5: Renaming Columns

**Your Task:**

Rename the `O3` column to "Ozone" and the `NO2` column to "Nitrogen_Dioxide" in the `united_data` dataset. Assign the result to a new variable named `renamed_data`.

```{r}
# Complete the code below and run the cell.

# Your code here
renamed_data <- 
```


#### Test Your Code (Exercise 5)

Do not modify the cell below. 
```{r}
library(testthat, quietly = TRUE)
test_that("Renaming columns", {
  expect_true("Ozone" %in% colnames(renamed_data))
  expect_true("Nitrogen_Dioxide" %in% colnames(renamed_data))
})

```

