---
title: 'Chapter 3: R Coding Basics'
output:
  html_document:
    df_print: paged
---

# Exercise 1: Variables and Assignments

**Your Task:**\
1. Define a variable called `my_age` and assign your age to it.\
2. Define another variable called `my_name` and assign your name to it.

**Consider:** What are the appropriate data types of `my_age` and `my_name`?

```{r}
# Complete the code and run the cell.

my_age <- # Your age here
my_name <- # Your name here

# Print the values
print(my_age)
print(my_name)
```

The `print()` function in R is used to display the values of variables or expressions on the screen. Unlike directly calling the variable, using `print()` provides explicitness and allows you to control the output format when displaying values.

### Test Your Code (Exercise 1)

Do not modify the cell below. Run the cell below to check if you've defined and assigned the variables correctly.

```{r}
# Load testthat package
library(testthat)

# Unittest for Exercise 1
test_that("Exercise 1: Variables and Assignments", {
  
  expect_that(my_age, is_a("numeric"))
  expect_gte(my_age, 0)
  expect_lte(my_age, 100)
  
  expect_that(my_name, is_a("character"))
  expect_gt(length(my_name), 0)
  expect_lte(length(my_name), 50)
  
})
```

------------------------------------------------------------------------

# Exercise 2: Installing Packages

**Your Task:**\
1. Install the `xgboost` package.\
2. Install `xml` and `mlr3` packages in one line of code by passing in two package names as a single variable.

**Consider:** You might want to run `?install.packages()` to see what data type it allows for the package names.

```{r}
# Complete the code below and run the cell.

# Install xgboost -- your code here
# Install xml and mlr3 -- your code here
```

**What are these packages?**\
`xgboost` is an efficient and scalable machine learning library designed for gradient boosting. It's widely used for supervised learning tasks, especially in structured/tabular data, due to its high performance and accuracy.

The `xml` package in R provides tools to parse and manipulate XML (eXtensible Markup Language) files, commonly used for data exchange between different systems and applications.

`mlr3` is an R package for machine learning in R that provides a unified interface and infrastructure for various machine learning tasks such as data preprocessing, model training, hyperparameter tuning, and evaluation. It's a successor to the popular mlr package and is designed to be more modular and extendable.

### Test Your Code (Exercise 2)

Do not modify the cell below. Run the cell below to check if you've successfully installed the packages.

```{r}
library(testthat)
test_that("Installing Packages", {
  # Task 1: Check if tidyverse is installed
  expect_true("xgboost" %in% installed.packages()[, "Package"])
  
  # Task 2: Check if ggplot2 and dplyr are installed
  packages_installed <- c("xml", "mlr3")
  expect_true(all(packages_installed %in% installed.packages()[, "Package"]))
})
```

------------------------------------------------------------------------

# Exercise 3: Basic Calculation

**Your Task:**\
1. Add 25 and 37, then divide it by 2. Assign it to a variable `A`.\
2. Multiply `A` by 5, then subtract it by pi(π). Assign it to a variable `B`.\
3. Raise `B` to the power of 5. Assign it to a variable "C".\
4. Test if C is an integer. Assign the test result to a variable "C_is_integer".

**Consider:** Operators in R: <https://www.w3schools.com/r/r_operators.asp>

```{r}
# Complete the code below and run the cell.

A <- # Your code here
B <- # Your code here
C <- # Your code here
C_is_integer <- # Your code here

print(C)
message(paste("Is C an integer?", C_is_integer))
```

Note that when we use `print(C)`, we get a value that looks like an integer. However, your `C_is_integer` result should say otherwise. This is resulted from the default behaviour of R that it rounds the number for display. Just remember not to be misled when using `print()` function!

### Test Your Code (Exercise 3)

Do not modify the cell below. Run the cell below to check if your calculation is correct.

```{r}
library(testthat)

# Unittest for Exercise 3
test_that("Exercise 3: Basic Calculation", {
  expect_equal(A, 31, tolerance=0)
  expect_equal(B, 151.8584, tolerance=0.0001)
  expect_equal(C, 80759608318.2073, tolerance=0.0001)
  expect_false(C_is_integer)
})
```

------------------------------------------------------------------------

# Exercise 4: Data Structures

**Your Task:**\
1. Create a numeric vector called `my_vector` with the numbers 3, 7, 2, and 9.\
2. Create a character vector called `my_names` with the names "Alice," "Bob," "Charlie," and "David."\
3. Create a list called `my_list` containing the vector `my_vector` and the vector `my_names`.\
4. Create a data frame called `my_dataframe` with two columns: "Names" and "Ages." Use the vectors `my_names` and `my_vector` as data for the columns.

**Consider:** Mastering the skill of searching online is crucial for programming. Now is the perfect opportunity to practice and enhance this skill!

```{r}
# Complete the code below and run the cell.

# 1. Create a numeric vector
my_vector <- # Your code here

# 2. Create a character vector
my_names <- # Your code here

# 3. Create a list
my_list <- # Your code here

# 4. Create a data frame
my_dataframe <- # Your code here
```

**Note:** There are numerous approaches to construct these data structures in R. For instance, we can create a data frame by directly inputting values, using other variables, or reading from a file. Additionally, you can explore various optional parameters to enhance your function. Embrace versatility and avoid limiting your solutions to just one method!

### Test Your Code (Exercise 4)

Do not modify the cell below. Run the cell below to check if your calculation is correct.

```{r}
library(testthat)

# Unittest for Exercise 4
test_that("Exercise 4: Data Structures", {
  expect_equal(my_vector, c(3, 7, 2, 9))
  expect_equal(my_names, c("Alice", "Bob", "Charlie", "David"))
  expect_true(is.list(my_list))
  expect_true(is.data.frame(my_dataframe))
  expect_equal(my_dataframe$Names, my_names)
  expect_equal(my_dataframe$Ages, my_vector)
})
```
