---
title: 'Chapter 9: Importing Your Data in R'
output:
  html_document:
    df_print: paged
---


# Getting Started:

Before diving into the exercises, follow the instructions below to download a 2020 winter (Toronto) data from the Exploring Air Quality Data website.

-   Visit the [Exploring Air Quality Data](https://uoft-chem.shinyapps.io/Air_Quality_App/) and go to `My Data` tab.
-   Enter your student number and then download the CSV file.
-   Upload the file to your RStudio. 


# Exercise 1: Importing data from CSV files

**Your Task:**

1. Use the `read_csv()` function from the `readr`(`tidyverse`) package to import your Air Quality dataset.  
2. Assign the imported data to a variable named `data`.

```{r}
# Complete the code below and run the cell.

# Import data from CSV file using read_csv -- your code here

```


#### Test Your Code (Exercise 1)

Do not modify the cell below. 

```{r}
library(testthat, quietly = TRUE)
test_that("Importing Data from CSV", {
  expect_true(is.data.frame(data) & nrow(data) > 0)
})
```

------------------------------------------------------------------------

# Exercise 2: Viewing Data

**Your Task:**

1. Display the first 5 rows of the `data` using an R built-in function we learned.  
2. Display the last 3 rows of the `data` using an R built-in function similar to the previous question.  

**Consider**: There is a sibling R built-in function to what you learned in the chapter. Look up what the function is called. 

```{r}
# Complete the code below and run the cell.

# View data -- your code here

```

**Questions:**

1. What are all the columns?   
2. What is the value of `O3` in the third row?  
3. What is the value of `NO2` of the last row?  

------------------------------------------------------------------------

# Exercise 3: Converting Tibble to Dataframe

**Your Task:**

1. Convert the `data` (which is a tibble) into a standard R dataframe.  
2. Assign this dataframe to a variable named `converted_data`.

```{r}
# Assume 'data' is the Tibble from the previous exercise

# Print to check the conversion
```

#### Test Your Code (Exercise 3)

Do not modify the cell below. 
```{r}
library(testthat, quietly = TRUE)
test_that("Converting Tibble to Dataframe", {
  expect_true(is.data.frame(converted_data))
})

```

------------------------------------------------------------------------

# Exercise 4: Saving Data

**Your Task:**

Save the `converted_data` as a new CSV file named "saved_data.csv" using the `write_csv()` function from `tidyverse`.  

Note: Save it to the current working directory. 


```{r}
# Complete the code below and run the cell.

# Save data as a CSV file -- your code here

```

#### Test Your Code (Exercise 4)

Do not modify the cell below. 
```{r}
library(testthat, quietly = TRUE)
test_that("Saving Data to Different Formats", {
  # Check if the CSV file exists
  expect_true(file.exists("saved_data.csv"))
})

```


------------------------------------------------------------------------

# (Optional) Exercise 5: Exploring what you can do with `read_csv()`

**Your Task:**. 
You are going to learn more about what you can do with `read_csv()`.

1. Use the `read_csv()` function to import only the first 10 rows of your original Air Quality data.  
3. Skip the first 2 rows of the data.  
3. Assign this data to a variable named `limited_data`.  

**Consider**: Take a look at the following tidyverse documentation: https://readr.tidyverse.org/reference/read_delim.html.  
-   The `n_max` parameter in `read_csv()` allows you to limit the number of lines/rows to read from the input file.
-   The `skip` parameter allows you to skip the first few rows of the file.

```{r}
# Complete the code below and run the cell.

# Import limited rows with specific parameters from CSV file -- your code here

```

#### Test Your Code (Exercise 5)

Do not modify the cell below. 

```{r}
library(testthat, quietly = TRUE)
test_that("Exploring Input Parameters of read_csv", {
  
  # Check if the number of rows in the imported data is 10
  expect_equal(nrow(limited_data), 10)
})
```


