---
title: 'Chapter 14: Summarizing Data'
output:
  html_document:
    df_print: paged
---

# Getting Started:

1. Navigate to the R4EnvChem-Exercises repository at https://github.com/UofTChem-Teaching/R4EnvChem-Exercises.
2. Inside the repository, go to the "Chapter 14" folder.
3. Look for the "ATR_plastics.csv" file in this folder and click on it to open the file.
4. On the top-right corner of the file view, click the three-dotted icon, where you'll find a "Download" button. Click on it to download the CSV file to your local machine.

## Load Necessary Libraries

Before starting the exercises, ensure you have the necessary libraries loaded:

```{r}
# Load necessary libraries
library(tidyverse)
library(flextable)
```

## Import Data

Use the `read_csv()` function from the `readr`(`tidyverse`) package to import your ATR Plastic dataset, and assign the imported data to a variable named `data`.

```{r}
# Complete the code below and run the cell.

# Import data from CSV file using read_csv -- your code here
data <- 
data
```


------------------------------------------------------------------------

# Exercise 1: Summarizing EPDM Data

In this exercise, we will work with the EPDM values.

**Your Task:**

Create a summary table named `summary_EPDM` by calculating the mean, median, minimum, maximum, and standard deviation of the `EPDM` values. Label these statistics as Mean_EPDM, Median_EPDM, Min_EPDM, Max_EPDM, and Std_Dev_EPDM, respectively. Then, make it into a flextable "summary_EPDM_ft".

```{r}
summary_EPDM <- # Your code here
  
summary_EPDM_ft <- flextable(summary_EPDM) # Modify this to make the table as pretty as you want

summary_EPDM_ft
```


#### Test Your Code (Exercise 1)

Do not modify the cell below. 

```{r,purl=FALSE}
library(testthat, quietly = TRUE)

test_that("Exercise 1: Summarizing EPDM Data", {
  expect_is(summary_EPDM_ft, "flextable")
  expect_true("Mean_EPDM" %in% colnames(summary_EPDM))
  expect_true("Median_EPDM" %in% colnames(summary_EPDM))
  expect_true("Min_EPDM" %in% colnames(summary_EPDM))
  expect_true("Max_EPDM" %in% colnames(summary_EPDM))
  expect_true("Std_Dev_EPDM" %in% colnames(summary_EPDM))
})
```

------------------------------------------------------------------------

# Exercise 2: Grouping and Summarizing by Rounded Wavenumber

In this exercise, you will group the data by rounded `wavenumber` values (rounded to the nearest 10s) and calculate various summary statistics for other columns. Finally, you will create a summary table using the `flextable()` function.

**Your Task:**

1. Round the `wavenumber` values to the nearest 10s (e.g., 1008.1060 and 1005.9078 will be rounded together).  
2. Group the data by the rounded `wavenumber` values.  
3. Calculate the mean and median of the `EPDM`, `Polystyrene`, and `Polyethylene` values within each group. Label these statistics as Mean_EPDM, Median_EPDM, Mean_Polystyrene, Median_Polystyrene, Mean_Polyethylene, and Median_Polyethylene, respectively.
4. Create a summary table using the `flextable()` function with appropriate headers and labels for the summary statistics.


```{r}
# Complete the code below and run the cell.

# Round the 'wavenumber' values to the nearest 10s
data_rounded <- 

# Group the data by the rounded 'wavenumber' values
data_grouped <- 

# Calculate summary statistics and create a flextable
summary_table <- 

  
flextable(summary_table) # Feel free to modify the flextable.
```


#### Test Your Code (Exercise 2)

Do not modify the cell below. 

```{r,purl=FALSE}
library(testthat, quietly = TRUE)

test_that("Grouping and Summarizing by Rounded Wavenumber", {
  expect_true("Mean_EPDM" %in% colnames(summary_table))
  expect_true("Median_EPDM" %in% colnames(summary_table))
  expect_true("Mean_Polystyrene" %in% colnames(summary_table))
  expect_true("Median_Polystyrene" %in% colnames(summary_table))
  expect_true("Mean_Polyethylene" %in% colnames(summary_table))
  expect_true("Median_Polyethylene" %in% colnames(summary_table))
})
```
