---
title: 'Chapter 4: R Coding Basics'
output:
  html_document:
    df_print: paged
---

# Exercise 1: Variables and Assignments

**Your Task:**

1. Define a variable called `my_age` and assign your age to it.  
2. Define another variable called `my_name` and assign your name to it.  

**Consider:** What are the appropriate data types of `my_age` and `my_name`?  

```{r}
# Complete the code and run the cell.

my_age <- # Your age here
my_name <- # Your name here

# Print the values
print(my_age)
print(my_name)
```

The `print()` function in R is used to display the values of variables or expressions on the screen. Unlike directly calling the variable, using `print()` provides explicitness and allows you to control the output format when displaying values.

#### Test Your Code (Exercise 1)

Do not modify the cell below. Run the cell below to check if you've defined and assigned the variables correctly.

```{r}
# Load testthat package
library(testthat)

# Unittest for Exercise 1
test_that("Exercise 1: Variables and Assignments", {
  
  expect_that(my_age, is_a("numeric"))
  expect_gte(my_age, 0)
  expect_lte(my_age, 100)
  
  expect_that(my_name, is_a("character"))
  expect_gt(length(my_name), 0)
  expect_lte(length(my_name), 50)
  
})
```

------------------------------------------------------------------------

# Exercise 2: Basic Calculation

**Your Task:**

1. Add 25 and 37, then divide it by 2. Assign it to a variable `a`.  
2. Multiply `a` by 5, then subtract it by pi(π). Assign it to a variable `b`.  
3. Raise `b` to the power of 5. Assign it to a variable "c".  
4. Test if `c` is an integer. Assign the Boolean result to a variable "c_is_integer".  

**Consider:** Operators in R: <https://www.w3schools.com/r/r_operators.asp>

```{r}
# Complete the code below and run the cell.

a <- # Your code here
b <- # Your code here
c <- # Your code here
c_is_integer <- # Your code here

print(c)
message(paste("Is c an integer?", c_is_integer))
```

Note that when we use `print(c)`, we get a value that looks like an integer. However, your `c_is_integer` result should say otherwise. This is resulted from the default behaviour of R that it rounds the number for display. Just remember not to be misled when using `print()` function!

#### Test Your Code (Exercise 2)

Do not modify the cell below. Run the cell below to check if your calculation is correct.

```{r}
library(testthat)

# Unittest for Exercise 2
test_that("Exercise 2: Basic Calculation", {
  expect_equal(a, 31, tolerance=0)
  expect_equal(b, 151.8584, tolerance=0.0001)
  expect_equal(c, 80759608318.2073, tolerance=0.0001)
  expect_false(c_is_integer)
})
```

------------------------------------------------------------------------

# Exercise 3: Data Structures

**Your Task:**

1. Create a numeric vector called `my_vector` with the numbers 3, 7, 2, and 9.  
2. Create a character vector called `my_names` with the names "Jessie," "Hyun," "Ali," and "Kevin."  
3. Create a list called `my_list` containing the vector `my_vector` and the vector `my_names`.  
4. Create a data frame called `my_dataframe` with two columns: "Names" and "Ages." Use the vectors `my_names` and `my_vector` as data for the columns.  


```{r}
# Complete the code below and run the cell.

# 1. Create a numeric vector
my_vector <- # Your code here

# 2. Create a character vector
my_names <- # Your code here

# 3. Create a list
my_list <- # Your code here

# 4. Create a data frame
my_dataframe <- # Your code here
```

**Note:** There are numerous approaches to construct these data structures in R. For instance, we can create a data frame by directly inputting values, using other variables, or reading from a file. Additionally, you can explore various optional parameters to enhance your function. Embrace versatility and avoid limiting your solutions to just one method!  

#### Test Your Code (Exercise 3)

Do not modify the cell below. Run the cell below to check if your code is correct.

```{r}
library(testthat)

# Unittest for Exercise 3
test_that("Exercise 3: Data Structures", {
  expect_equal(my_vector, c(3, 7, 2, 9))
  expect_equal(my_names, c("Jessie", "Hyun", "Ali", "Kevin"))
  expect_true(is.list(my_list))
  expect_true(is.data.frame(my_dataframe))
  expect_equal(my_dataframe$Names, my_names)
  expect_equal(my_dataframe$Ages, my_vector)
})
```

------------------------------------------------------------------------

# Exercise 4: Calculate Median

**Your Task:**

1. Create a numeric vector called `nums` containing any ten numbers of your choice.  
2. Calculate the median of `nums` and store the result in a variable called `median_value`.

**Consider:** What R built-in function can you use here? You can search online.

```{r}
# Complete the code below and run the cell.

# 1. Create a numeric vector
nums <- # Your code here

# 2. Calculate the median
median_value <- # Your code here
```

#### Test Your Code (Exercise 4)

Do not modify the cell below. Run the cell below to check if your code is correct.

```{r}
library(testthat)

# Do not use this function to solve this question!
med <- function(numbers) { 
  sorted_numbers <- sort(numbers)
  n <- length(sorted_numbers)
  if (n %% 2 == 0) {
    return((sorted_numbers[n/2] + sorted_numbers[n/2 + 1]) / 2)
  } else {
    return(sorted_numbers[(n + 1) / 2])
  }
}

# Unittest for Exercise 4
test_that("Exercise 4: Calculate Median", {
  expect_true(length(nums) == 10)
  expect_equal(median_value, med(nums))
})
```


------------------------------------------------------------------------

# Exercise 5: Conditional Statement

**Your Task:**

Given a variable `grade` that represents a numeric grade (0 to 100), write an if-else if-else structure to assign a letter grade based on the following criteria:

-   "F" if the grade is less than 50
-   "D" if the grade is between 50 and 59
-   "C" if the grade is between 60 and 69
-   "B" if the grade is between 70 and 79
-   "A" if the grade is 80 or above.

Your solution should store the letter grade in a variable named `letter_grade`.

**Consider:** When there are multiple criteria, you can use `else if` syntax. Check out https://www.r-bloggers.com/2022/10/if-else-else-if-statement-in-r/ for examples.


```{r}
# Complete the code below and run the cell.
grade <- 75  # Change this value to test different scenarios

# Your code goes here



# Is your output what you expect?
print(letter_grade)
```

